--- 
title: "Environmental Systems Data Science"
author: "Loïc Pellissier, Joshua Payne, Benjamin Stocker"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook: default
#  pdf_document:
#    latex_engine: xelatex
bibliography: [book.bib] # Packages.bib excluded
biblio-style: apalike
link-citations: true
nocite: '@*'
description: "Text book and exercises for the course Environmental System Data Science at ETH Zürich."
---

# Prerequisites {-}

## Course Description {-}
Students are introduced to a typical data science workflow using various examples from environmental systems. They learn common methods and key aspects for each step through practical application. The course enables students to plan their own data science project in their specialization and to acquire more domain-specific methods independently or in further courses.

## Course Objectives {-}
The students are able to

- frame a data science problem and build a hypothesis
- describe the steps of a typical data science project workflow
- conduct selected steps of a workflow on specifically prepared datasets, with a focus on choosing, fitting and evaluating appropriate algorithms and models
- critically think about the limits and implications of a method
- visualise data and results throughout the workflow
- access online resources to keep up with the latest data science methodology and deepen their understanding

## Content {-}
- The data science workflow
- Access and handle (large) datasets
- Prepare and clean data
- Analysis: data exploratory steps
- Analysis: machine learning and computational methods
- Evaluate results and analyse uncertainty
- Visualisation and communication

## Useful Prerequisites {-}
- 252-0840-02L Anwendungsnahes Programmieren mit Python
- 401-0624-00L Mathematik IV: Statistik
- 401-6215-00L Using R for Data Analysis and Graphics (Part I)
- 401-6217-00L Using R for Data Analysis and Graphics (Part II)
- 701-0105-00L Mathematik VI: Angewandte Statistik für Umweltnaturwissenschaften

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r include=FALSE}
# Preferences for conflicting packages
library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("levelplot", "rasterVis")
conflict_prefer("origin", "raster")
conflict_prefer("extract", "raster")
conflict_prefer("partial", "pdp")
conflict_prefer("mse", "Metrics")
```

```{r global-options, include=FALSE}
knitr::opts_chunk$set(out.width = "50%", fig.align = "center", warning=FALSE, message=FALSE)
options(tinytex.verbose = TRUE)
```
