---
title: "Exercise 02 - Solution"
output: html_document
---

# Data Wrangling

**1. Identify outlier:** Based on the half-hourly dataset for site CH-Lae, aggregated to daily means, identify outliers in `GPP_NT_VUT_REF` with respect to the linear relationship between `GPP_NT_VUT_REF` and `PPFD_IN`. To do so, first fit a linear regression model using `lm()`. This function returns a list of objects, one of which is `residuals`. Determine outliers as the "outlying" points in the distribution of residuals. You may use the base-R function `boxplot.stats()` and set the argument `coef` accordingly to our customised threshold definition.

```{r}
## 1. Identify Outlier
# Load packages
library(tidyverse) 
library(lubridate)  # not part of the automatic load of tidyverse

# read half-hourly data
hhdf <- read_csv("../data/FLX_CH-Lae_FLUXNET2015_FULLSET_HH_2004-2014_1-3.csv") %>% 
  
    # interpret -9999 as missing value
    na_if(-9999) %>% 
  
    # interpret timestamp as a date-time object
    mutate_at(vars(starts_with("TIMESTAMP_")), ymd_hm)


# aggregate to daily
ddf <- hhdf %>% 
  
    # create a date-only object
    mutate(date = as_date(TIMESTAMP_START)) %>% 
  
    # aggregate
    group_by(date) %>% 
    summarise(GPP_NT_VUT_REF = mean(GPP_NT_VUT_REF, na.rm = TRUE),
              PPFD_IN = mean(PPFD_IN, na.rm = TRUE),
              .groups = 'drop' # not mandatory
    )

# fit linear regression
linmod <- lm(GPP_NT_VUT_REF ~ PPFD_IN, data = ddf)

# get box plot statistics for determining "outlying" points
out_boxplot <- boxplot.stats(linmod$residuals)

# record the row numbers of outlying points based on the output list element 'out'
# row numbers are the names of elements in out_boxplot$out, provided as strings. 
# convert them to integers.
idx_outlying <- names(out_boxplot$out) %>% as.integer()

```

***
**2. Remove outliers by setting values in the data frame (aggregated daily data frame for CH-Lae) to `NA`.**
```{r}
## 2. Remove Outliers
# In base-R, this could be done as:
ddf$GPP_NT_VUT_REF_clean <- ddf$GPP_NT_VUT_REF
ddf$GPP_NT_VUT_REF_clean[idx_outlying] <- NA

# In tidyverse style:
ddf <- ddf %>% 
  mutate(rownumber = row_number()) %>%    # could also do: mutate(rownumber = 1:nrow(.))
  mutate(GPP_NT_VUT_REF_clean = ifelse(rownumber %in% idx_outlying, NA, GPP_NT_VUT_REF))
```

***
**3. Create a scatterplot of all daily data (GPP vs. PPFD) and highlight outliers that are removed by step 2.**
```{r}
## 3. Create scatterplot
ddf %>% 
  ggplot(aes(x = PPFD_IN)) + 
  geom_point(aes(y = GPP_NT_VUT_REF), color = "red") +  # first, plot all points including outliers in one color
  geom_point(aes(y = GPP_NT_VUT_REF_clean), color = "black")  # then overplot cleaned data in black so that remaining red points are outliers
```

***
**4. Visualising diurnal and seasonal cycles:** Using the half-hourly dataset for site CH-Lae, visualise how GPP (`GPP_NT_VUT_REF`) varies on two time scales: diurnal (within-day at hourly time scale) and seasonal.
To implement this, follow the following steps:

a. Summarise half-hourly data for each data across multiple years to get a mean seasonality with a mean diurnal cycle for each day of the year. You will use functions from the lubridate package (e.g., `yday()`). To deal with date-time objects, use the lubridate package. Enter `?day` to get more hints.
b.  Create a raster plot (`geom_raster()`), mapping the hour of the day to the x-axis, the day of the year to the y-axis, and the magnitude of `GPP_NT_VUT_REF` to color (fill).
c.  Make this figure ready for publication by adding nice labels and choosing a good color scale.
    
```{r}
## 4. Visualising cycles
## a. Half-hourly dataset
hhdf_meanseason <- hhdf %>% 
  mutate(hour_day = hour(TIMESTAMP_START),
         day_year = yday(TIMESTAMP_START)) %>% 
  group_by(hour_day, day_year) %>% 
  summarise(gpp = mean(GPP_NT_VUT_REF, na.rm = TRUE))

## b. Raster plot
hhdf_meanseason %>% 
  ggplot(aes(x = hour_day, y = day_year, fill = gpp)) +
  geom_raster()

## c. Make raster plot publishable - This is up to your judgement. Below is an example. 
library(viridisLite)

hhdf_meanseason %>% 
  
  # as above
  ggplot(aes(x = hour_day, y = day_year, fill = gpp)) +
  geom_raster() +
  
  # use a color scale that works also for color-blind people
  scale_fill_viridis_c(option = "magma") +
  
  ## change theme (axes style)
  theme_classic() +
  
  # adjust the aspect ratio of the plotting region
  coord_fixed(ratio=0.2) +
  
  # labels of each mapping axis, \n is a line break
  labs(x = "Hour of day", y = "Day of year", fill = expression(paste("GPP \ngC m"^-2, "s"^-1))) +
  
  # avoid adding distance from the lowest values to the axes
  scale_x_continuous(expand = c(0,0)) + scale_y_continuous(expand = c(0,0))
```
